\bf Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid

### 2015-02-17 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/main.tex
Part of the response to reviewers

Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
This method uses a pyramid structure of populations to combine model based search with local search
to achieve parameter-less optimization. Initial results suggest that unlike
previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
parameterized search algorithms. In this work we shall: extend these results to cover more
comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
fitnesses; algorithm complexity analysis; and provide more in depth analysis of P3 itself.

@@

This goal is formalized using model scoring from Bayesian statistics. In its raw form this
almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
derived a simplified form shown in Equation~\ref{eq-hboa-final}. If no proposed split satisfies the inequality, iteration stops.
If multiple splits do, whichever maximizes the right side is chosen.

@@

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$
      \State Add solution to $hashset$
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$
      \If{solution's fitness has improved}
        \If{solution $\notin hashset$}
          \State Add solution to $P_{i+1}$
          \State Add solution to $hashset$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization. $pyramid$ is an
           ordered set of populations and $hashset$ is a set
           of all solutions in $pyramid$.}
  \label{fig-p3}
\end{figure}

@@

\subsection{Quantitative Comparison}
Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.

@@

\subsubsection{Model Building}
hBOA and Parameter-less hBOA perform significantly worse when using wall clock time as the unit of comparison than when using
evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
hBOA goes from being slightly more efficient than LTGA in terms of evaluations to significantly worse in terms of seconds.
As P3 and LTGA require a similar asymptotic complexity per
evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.

@@

\section{Population Sizing}
A major advantage to P3 is that it does not require the user to set a population size parameter.
Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
to initialization and no need to sacrifice intermediate fitness for eventual optimality.

@@

\begin{figure}
  \begin{center}
  \includegraphicsfit{evals-to-success-range}
  \end{center}
  \caption{Comparison of the upper and lower quartiles of evaluations required
           to reach the global optimum for P3 and LTGA with respect to problem size.}
  \label{fig-evals-to-success-range}
\end{figure}

>>>>

Most recently \cite{goldman:2014:p3} introduced the Parameter-less Population Pyramid~(P3).
This method uses a pyramid structure of populations to combine model based search with local search
to achieve parameter-less optimization. Initial results suggest that unlike
previous parameter-less methods, P3 is actually more efficient than current state-of-the-art
parameterized search algorithms. In this work we shall: extend these results to cover more
comparison algorithms; compare both efficiency in reaching the global optimum and intermediate
fitnesses; analyze algorithm complexity; and provide more in depth analysis of P3 itself.

@@

This goal is formalized using model scoring from Bayesian statistics. In its raw form this
almost always creates near infinitesimal results, calculating fractions which include factorials of $\mu$
and products over $N$. However, through algebraic manipulation discussed in Appendix~\ref{sec-appendix}, we
derived a simplified form shown in Equation~\ref{eq-hboa-final}.
Here $l$ is a leaf in tree $i$, with $l'$ and $l''$ the results of splitting $l$. $m_i(l)$ is the number
of solutions which reach $l$ and $m_i(x_i, l)$ is the number of solutions which reach $l$ with the given value
for $x_i$.
If no proposed split satisfies the inequality, iteration stops.
If multiple splits do, whichever maximizes the right side is chosen.

@@

\begin{figure}
  \begin{algorithmic}[1]
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber
    \If{solution $\notin hashset$}
      \State Add solution to $P_0$ and rebuild model
      \State Add solution to $hashset$
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$
      \If{solution's fitness has improved}
        \If{solution $\notin hashset$}
          \State Add solution to $P_{i+1}$ and rebuild model
          \State Add solution to $hashset$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization. $pyramid$ is an
           ordered set of populations and $hashset$ is a set
           of all solutions in $pyramid$.}
  \label{fig-p3}
\end{figure}

@@

\subsection{Quantitative Comparison}
Of the 130 tested configurations, P3 found the global optimum using the least median evaluations on 114.
The largest problem size for any problem where P3 was not the most efficient has 49 bits, with P3 achieving the best results on all 92 larger
configurations. hBOA, LTGA, and Parameter-less hBOA only outperform P3 on the smallest 5, 4, and 1 Deceptive
Step Trap instances, respectively. Random Restart Hill Climbing outperforms P3 on the smallest 3 Nearest Neighbor NK instances
and the smallest Ising Spin Glass. $1+(\lambda, \lambda)$ has the most success outperforming P3, doing so
on the smallest 5 Deceptive Traps, 3 smallest Deceptive Step Traps, and 2 smallest Rastrigin.
Pairwise comparison of LTGA and P3 on the largest problem size using the Mann-Whitney~U test results in
$p < 10^{-5}$ for all problems.

@@

\subsubsection{Model Building}
hBOA and Parameter-less hBOA perform much worse when using wall clock time as the unit of comparison than when using
evaluations. This makes sense as hBOA's model building requires $\Omega(N^2)$ time per evaluation while, under reasonable
assumptions, P3 and LTGA require $O(N)$ time per evaluation. This penalty is most clear on Ising Spin Glass where
hBOA goes from being slightly more efficient than LTGA in terms of evaluations to three orders of magnitude worse in terms of seconds.
As P3 and LTGA require a similar asymptotic complexity per
evaluation as the Hill Climber and $1+(\lambda, \lambda)$, no similar change in ordering occurs.

@@

\section{Population Sizing}
\begin{figure}[t]
  \begin{centering}
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{pop-p3}
      \end{centering}
      \caption{As problem size increases}
      \label{fig-pop-p3}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \begin{centering}
        \includegraphicsfit{solutions-stored}
      \end{centering}
      \caption{Largest problem size}
      \label{fig-solutions-stored}
    \end{subfigure}
  \end{centering}
  \caption{The total number of solutions stored by P3 when the global optimum is found.
           In Figure~\ref{fig-solutions-stored} the red ``+'' indicates LTGA's tuned population size.}
  \label{fig-p3-storage}
\end{figure}

A major advantage to P3 is that it does not require the user to set a population size parameter.
Beyond making P3 easier to apply, this also conveys two additional advantages: diversity scaled
to initialization and no need to sacrifice intermediate fitness for eventual optimality.

@@

\begin{figure}
  \begin{center}
  \includegraphicsfit{evals-to-success-boxplot}
  \end{center}
  \caption{Comparison of the upper and lower quartiles of evaluations required
           to reach the global optimum for P3 and LTGA with respect to problem size.}
  \label{fig-evals-to-success-boxplot}
\end{figure}

### 2015-02-17 Fast and Efficient Black Box Optimization using the Parameter-less Population Pyramid/reviews.tex
Part of the response to reviewers

We have explained that, for the particular situation in the bottom pyramid level, before a randomly generated solution is added to that level, that solution should be hill climbed. Our reasoning is straightforward; a randomly generated solution is unlikely to have any useful linkage information to be incorporated into that level. Once hill-climbed, that solution is more likely to provide useful linkage information to that bottom level. There is no guarantee that \textbf{every} such hill-climbed solution will be improved, and to the extent that it is not the erroneous linkage information added at the bottom level will need to be repaired at higher levels.  The bottom line is that local search, in this particular role, is a fundamental aspect of P3 and P3's performance as we see it.

@@

Our job, as algorithm designers, is to compare our approach against the best possible competing algorithms and report how our approach fares. We have taken, to the best of our ability using available published resources, those best algorithms and  P3 and compare them each against a rigourous set of problems. In fact, we feel we went quite a bit out of our way to \textbf{tune} these algorithms so that we may compare on the best possible footing, in other words to be fair.

Our jobs is not to speculate on possible changes to these algorithms and how, given those changes, the ``new and improved'' algorithms might perform. If subsequent researchers wish to do so, we welcome their work. In fact, we hope that this paper inspires others to go in new directions (paramaterless algorithms, local search integration regardless of problem, etc.)

>>>>

We have explained that, for the particular situation in the bottom pyramid level,
before a randomly generated solution is added to that level, that solution should be hill climbed.
Our reasoning is straightforward; a randomly generated solution is unlikely to
have any useful linkage information to be incorporated into that level.
Once hill-climbed, that solution is more likely to provide useful linkage information to that bottom level.
This is true even for problems with large fitness plateaus, such as Deceptive Step Trap.
As such we feel the application of local search should be considered the standard approach
with P3, only to be removed under exceptional circumstances.

@@

Our job, as algorithm designers, is to compare our approach against the best possible competing algorithms
and report how our approach fares. We have taken, to the best of our ability using available published resources,
those best algorithms and compared them with P3 on a rigorous set of problems.
We went out of our way to tune these algorithms
so that we may compare on the best possible footing, giving them an advantage not available to P3.

However, adding initial local search as a tunable parameter to hboa and LTGA when partial evaluations are not available is not part
of existing work for either method. Doing so would represent novel research outside the scope of our topic, and provide those techniques
with even more problem specific configuration. We welcome future work by other researchers exploring the possible effects of combining
local search with their techniques.
